### 2020-08-26 thesis/chap_2.tex
re
update discussion chap 2

This research is an extension of the work intitiated by the project “Monteregie Connection”, whose goal was to engage stakeholders around a discussion of the ecosystem services provided by natural habitat in the region. In the context of this stakeholder-driven project, \cite{mitchell_monteregie_2015} conducted a non-spatially explicit scenario building exercise for a subset of the Monteregie region, the Vallée-de-Richelieu Municipalité Region Comté. The outcome was 4 scenarios, ranging from “Periurban Development” to “Green Development”, and each describing different land use change trends. The Monteregie connection project improved our understanding of the relationship between forest connectivity and ecosystem services. However, there was no effort to estimate the impact of each scenario on ecological connectivity. This is partially due to a lack of tools able to translate these scenarios into simulations. We contribute to the development of such tools by integrating the results of the workshop into the land use change model developed in chapter 1. The goal of this chapter is therefore twofold: gain a better understanding of conservation priorities in the region, and to provide a basis for an iterative scenario-building process.

\\
%-----------------------------------------------

@@

Conducting participatory research in Monteregie led us to a better understanding of the obstacles and assets to connectivity conservation in the region. The predominance of the agricultural land cover and its implications for land use planning and the regional economy was revealed as the most important source of obstacles for the region’s conservation goals, but also as a source of opportunities to develop more biodiversity-friendly farming practices. The expansion of urban infrastructure ranked high in more forested regions of the landscape. A range of assets was identified in the region: the strong engagement of local actors in connectivity conservation was a key factor. Through a consensual mapping workshop, we also mapped consensual priority core areas and links, showing how participants think corridors should make the best use of linear strips of the remaining forest in the region. Specifically, this demonstrates that participants saw small patches as important for corridor design and perceived corridors as an important tool to conserve and restore the potential of fragmented patches. However, it also means that designing corridors in such a way might mean that an otherwise important patch could be ignored because its location does not fit a linear corridor design.

Interestingly, Different tables also identified corridors with variable redundancy. Redundancy is an important connectivity conservation design, increasingly being explored in connectivity experiments \citep{fletcher_matrix_2014}. It is possible that participants at the north table though redundancy was necessary to make the best use of the layout of forest, which are interspersed in the agricultural matrix. This is by opposition to the corridors of the south table, where forest patches are smaller and more scarce. Participants at this table might have thought that one non redundant linear corridor was all they could ask for in a predominant agricultural matrix. Those results represent a first step toward a better model of how stakeholders could make complex conservation-related decisions.

The integration of those corridors to the land use change model shows how qualitative priorities from a participatory process can be turned into model inputs. It is interesting to see that some changes anticipated by workshop participants are observed in our land use change business as usual scenario. For instance, the increasing land development in the east of the region toward estrie, coined as “villegiature pressure” by participants to designate the building of cottages and country houses. This demonstrates that participants are in tune with current land use change in the region and that our model was able to replicate them. Concerning connectivity results, it is important to see how they compare to the business as usual scenario. These results complement those of chapter 1 and show similar patterns with regards to species response. When only looking at mean flow values, species with linear and simple response are those who seem to depend on forest types that are only under threat by land use change. On the other hand, species dependent on forest types that are under threat both by land use change and climate change show a more complex response, with more variations in the final outcome (2100 horizon) of the scenarios.

Concerning flow complexity, it is interesting to note that the number of features detected by the SURF algorithm is consistently lower through time for all species under targeted reforestation (in corridors) - except for the Marten. It is possible that corridors allow to streamline current flow and prevents complexification that arises with fragmentation, although, as pointed out in chapter 1, the link between number of features and and complexity would need to be examined in more details before drawing a final conclusion.

We extend on the work of \cite{albert_applying_2017}, which implemented a simple land use change model, but did not attempt to generate any alternative scenarios. In many ways, this work can be compared to recent work that looked at how connectivity could change under a range of scenarios. One example is \cite{rubio_sustaining_2012} who tested whether a network of forested patches resilient to different land use change scenarios, for a species of nuthatch. They found that although identifying such a network was possible, its effectiveness and efficiency was under threat under certain scenarios. Our results suggest that findings of that sort are likely to change when climate change is considered, if the species under study is dependent on sensitive forest types.

The results of our simple scenarios have implications for connectivity conservation in the region. The reforestation scenario suggests that reforesting as much land as is urbanized is what is necessary to maintain mean current flow. But this is only one simple measure of connectivity, and therefore those results suggest that more work is needed to explore which metric would be most appropriate to look at in this context, and more importantly what land use practices (protection, reforestation, urban densification) would be enough to maintain this metric. In the light of the results from our workshop, it is clear that reforesting the same amount that we are urbanizing is possibly socially unacceptable, given the amount of agricultural loss it generates. Even such a simple scenario echoes the long debated question of land sparing versus land sharing, a dichotomy that is now largely questioned as it fails to address the complexity of both food systems and the diversity of farming practices \cite{tscharntke_global_2012, bennett_changing_2017}. In fact our reforestation scenario can be seen as an agricultural intensification model, in which farmers are able to cultivate less land to allow for more space for forest. Yet our workshop results suggest that stakeholders would not think along those lines, and are more likely to propose compensations to farmers for the loss in productive surface or for changes in practices. This is likely reinforced by the fact that some stakeholders are opposed to further intensification of agricultural practices (also cited as an obstacle by participants). New compensation schemes such as ALUS (Alternative Land Use Services) are being experimented in Monteregie and are looking promising in their capacity to federare a bottom up movement of conservation in agrosystems \cite{ouellet_community_2020}.

@@

Our methodology would allow us to explore alternative scenarios with ease. We imagined simple scenarios and therefore served as an intermediary between the workshop output and the model inputs. However, given the flexibility of the STSIM land use change model framework, steps could be taken to reduce that distance between stakeholders outputs and model inputs. The interest of stakeholders in participating in our workshop demonstrates that local actors are longing for tools that allow them to close the loop between what they think should be done and what it would effectively mean for species on the ground, to explore where and how much reforestation would provide the optimal corridor design, or how much redundancy is really needed. To make this a reality, we are missing some knowledge about specific parameters - for instance, what the ideal width of a corridor for a given species, or how much agricultural loss is acceptable to reforestation. But we are also missing specific tools, and this project demonstrates that it is now easier to get there. The STSIM framework has been used in a participatory setting before. For instance, in \cite{jarnevich_developing_2019}, STSIM was used to produce a model for invasive species spread and fire management, using expert knowledge collected in a participatory setting. The authors also highlight the value to be found in involving “end users” of model results (i.e. stakeholders) in the modelling process. There remains a lot of work to be done to put these analyses in the hands of stakeholders. Making the analysis open and the code reproducible is one step in this direction.

Perhaps the most important limitation of this work is the fact that the strong consensus which was quickly built during the workshop is fragilized by the absence of some key stakeholders which, in spite of being invited, were not able to partake in the activity. This further illustrates the need for consensual mapping and the need for an iterative process with the goal to involve a larger diversity of stakeholders. Participatory research scholars have discussed the importance to consult and involve a diverse set of participants, representative of different knowledge systems, in efforts of participatory modelling \cite{gray_modeling_2012}. The lack of representatives from the forestry and farming industries, important stakeholders in forest management, limits the impact of consensus building.

@@

Through participatory research we demonstrated the value of stakeholder engagement to models of connectivity conservation planning. We contributed a better understanding of obstacles to connectivity conservation and highlighted the flexibility of a scenario approach for the consideration of alternative scenarios of land use planning and connectivity conservation. Under our community-based scenarios, we observed similar types of response (linear versus nonlinear, simple versus complex) in our study species. Although we provided a basis for an iterative framework of community-driven scenario building, future research will need to achieve better representation in stakeholder groups.

>>>>

This research is an extension of the work intitiated by the project “Monteregie Connection”, whose goal was to engage stakeholders around a discussion of the ecosystem services provided by natural habitat in the region. In the context of this stakeholder-driven project, \cite{mitchell_monteregie_2015} conducted a non-spatially explicit scenario building exercise for a subset of the Monteregie region, the Vallée-de-Richelieu Municipalité Region Comté. The outcome was 4 scenarios, ranging from “Periurban Development” to “Green Development”, and each describing different land use change trends. The Monteregie connection project improved our understanding of the relationship between forest connectivity and ecosystem services. However, there was no effort to estimate the impact of each scenario on ecological connectivity. This is partially due to a lack of tools able to translate these scenarios into simulations. We contribute to the development of such tools by integrating the results of the workshop into the land use change model developed in chapter 1. The goal of this chapter is therefore twofold: gain a better understanding of conservation priorities in the region, and to provide a basis for an iterative scenario-building process. \\
%-----------------------------------------------

@@

Conducting participatory research in Monteregie led us to a better understanding of the obstacles and opportunities for connectivity conservation in the region. The predominance of the agricultural land cover and its implications for land use planning and the regional economy was revealed as the most important obstacle for the region’s conservation goals, but was also seen as an opportunity to develop more biodiversity-friendly farming practices. The expansion of urban infrastructure also ranked high in more forested regions of the landscape. A range of assets was identified in the region: the strong engagement of local actors in connectivity conservation was a key factor. Through a consensus mapping workshop, we also mapped consensus priority core areas and links, showing how participants think corridors would make the best use of linear strips of the remaining forest in the region. Specifically, this demonstrated that participants saw small patches as important for corridor design and perceived corridors as an important tool to conserve and restore the potential of fragmented patches.

The tendency toward linear design for participants could indicate that creating corridors in such a way might lead to situations where an otherwise important patch could be ignored because its location does not fit a linear corridor design. The relationship between linearity and patch importance for connectivity, and how it is perceived by stakeholders, is an important avenue of research for connectivity conservation. For example, it would be interesting to know which design is more socially acceptable versus the most impactful for connectivity restoration.

While linearity seemed important to all tables, some participants had different conceptions of the optimal corridor design for their subregions. For example different tables also identified corridors with variable redundancy. Redundancy is an important connectivity conservation design, increasingly being explored in connectivity experiments \citep{fletcher_matrix_2014}. It is possible that participants at the north table thought redundancy was necessary to make the best use of the layout of forest, which is interspersed in the agricultural matrix. This is by opposition to the corridors of the south table, where forest patches were smaller and more scarce. Participants at this table might have thought that one non redundant linear corridor was all they could ask for in a predominant agricultural matrix. Those results represent a first step toward a better model of how stakeholders might make complex conservation-related decisions.

The integration of the corridors identified by the participants into the land use change model shows how qualitative priorities from a participatory process can be turned into model inputs. It is interesting to see that some changes anticipated by workshop participants are observed in our land use change business-as-usual scenario. For instance, the increasing land development in the east of the region toward Estrie; this was coined as “villegiature pressure” by participants to designate the building of cottages and country houses. This demonstrates that participants are in tune with current land use change in the region and that our model was able to replicate them. Concerning connectivity results, it is important to see how they compare to the business as usual scenario. These results complement those of chapter 1 and show similar patterns with regards to species’ responses. When only looking at mean flow values, species with linear and simple responses are those that seem to depend on forest types that are only under threat by land use change. On the other hand, species dependent on forest types that are under threat from land use change and climate change show a more complex response, with more variations in the final outcome (2100 horizon) of the scenarios.

Concerning flow complexity, it is interesting to note that the number of features detected by the SURF algorithm is consistently lower through time for all species under targeted reforestation (in corridors) - except for the American marten. It is possible that corridors allow a streamlining of current flow and prevent complexification (i.e. reticulation) of flow that arises with fragmentation. However, as pointed out in chapter 1, the link between number of features and complexity would need to be examined in more detail before drawing a final conclusion on this observation.

We extended the work of \cite{albert_applying_2017}, which implemented a simple land use change model, but did not attempt to generate any alternative scenarios. In many ways, this work can be compared to recent work that looked at how connectivity might change under a range of scenarios. One example is \cite{rubio_sustaining_2012} who tested whether a network of forested patches was resilient to different land use change scenarios, for the european nuthatch (\textit{Sitta europaea}). They found that although identifying such a network was possible, its effectiveness and efficiency was under threat under certain scenarios. Our results suggest that findings of that sort are likely to differ when climate change is considered, if the species under study is dependent on climate-sensitive forest types.

The results of our simple scenarios have implications for connectivity conservation in the region. The reforestation scenario suggests that reforesting as much land as is urbanized is what is necessary to maintain mean current flow. But this is only one simple measure of connectivity, and therefore those results suggest that more work is needed to explore which metrics would be most appropriate to look at to support connectivity conservation planning, and more importantly what land use practices (protection, reforestation, urban densification) would be required to support the management of the ecological properties measured by this metric. In the light of the results from our workshop, it is clear that reforesting the same amount that we are urbanizing is possibly socially and economically unacceptable, given the amount of agricultural loss it generates. Even such a simple scenario echoes the long debated question of land sparing versus land sharing, a dichotomy that is now largely questioned as it fails to address the complexity of both food systems and the diversity of farming practices \cite{tscharntke_global_2012, bennett_changing_2017}. In fact, our reforestation scenario can be seen as an agricultural intensification model, in which farmers are able to cultivate less land to allow for more space for forest. Our workshop results suggest that stakeholders are instead likely to propose compensations to farmers for the loss in productive surface or for changes in practices (as was mentioned at some tables). This is likely reinforced by the fact that some stakeholders are opposed to further intensification of agricultural practices (also cited as an obstacle by participants). New compensation schemes such as ALUS (Alternative Land Use Services) are being experimented in Monteregie and are looking promising in their capacity to federare a bottom up movement of conservation in agrosystems \cite{ouellet_community_2020}.

@@

Our methodology would allow us to explore alternative scenarios with ease. We imagined simple scenarios and therefore served as an intermediary between the workshop output and the model inputs. However, given the flexibility of the STSIM land use change model framework, steps could be taken to reduce that distance between stakeholders outputs and model inputs. The interest shown by the  stakeholders to participate in our workshop demonstrates that local actors are longing for tools that allow them to close the loop between what they think should be done and what it would effectively mean for species on the ground. Models can help explore where and how much reforestation would provide the optimal corridor design, or how much corridor redundancy is really needed. To make this a reality, we are missing some knowledge about specific parameters - for instance, what the ideal width of a corridor is for a given species, or how much loss of agricultural land is acceptable to reforestation. But we are also missing specific tools, and this project demonstrates that it is now easier to get there. The STSIM framework has been used in a participatory setting before. For instance, in \cite{jarnevich_developing_2019}, STSIM was used to produce a model for invasive species spread and fire management, using expert knowledge collected in a participatory setting. The authors also highlighted the value to be found in involving “end users” of model results (i.e. stakeholders) in the modelling process. There remains a lot of work to be done to put these analyses in the hands of stakeholders. Making the analysis open and the code reproducible is one step in this direction which we have done for the chapters of this thesis.

Perhaps the most important limitation of this work is the fact that the strong consensus which was quickly built during the workshop occurred in the absence of representatives from the forestry and farming industries which limits the representativity of consensus that was built. The participants present were mainly from NGOs, regional land use planning offices and ministries. This further illustrates the need for consensus mapping and the need for an iterative process with the goal to involve a larger diversity of stakeholders. Participatory research scholars have discussed the importance to consult and involve a diverse set of participants, representative of different knowledge systems, in efforts of participatory modelling \cite{gray_modeling_2012}.

@@

Through participatory research we demonstrated the value of stakeholder engagement to models of connectivity conservation planning. We contributed a better understanding of the obstacles to connectivity conservation and highlighted the flexibility of a scenario approach for the consideration of alternative scenarios of land use planning and connectivity conservation. Under our community-based scenarios, we observed similar types of response (linear versus nonlinear, simple versus complex) in our study species. Although we provided a basis for an iterative framework of community-driven scenario building, future research will need to achieve better representation in stakeholder groups.
